Local Search Approximation Algorithms for Partitional and Hierarchical Clustering

### 2015-05-10 tex/project.tex
Further updated bib and down to 8 pages!

\begin{figure}[h]
\begin{tabular}{ | l | l | l | l | l | }
\hline
	Data Set Name &Optimal Objective & $k$ & $n$ & dim   \\ \hline
	Soybean & 113.8 & 4 & 47 & 35 \\ 
	Iris & 98.2 & 3 & 150 & 4 \\ 
	Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
	Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
	Gauss\_5\_50\_5 & 178.1 & 50 & 408 & 5 \\ 
	pmed1 & 5819 & 5 & 100 & \  \\ 
	\hline
\end{tabular}
\caption{Some specifics about our datasets}
\end{figure}

Tests were conducted using Python 2.7.9 on a 64-bit Windows platform, on a 4 core Intel processor.

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	pmed1 & N/A & N/A & 0.003 & 5.042 & 0.084 & 0.09 & 0.155 & 0.161 & 0.255 \\ \hline
\end{tabular}

\caption{Running Times For Algorithms}
\end{figure*}

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline
	Soybean & 0.69 & 0.18 & 0.55 & 0.61 & 0.71 & 0.73 & 0.72 & 0.71 & 0.71  \\ 
	Iris & 0.75 & 0.76 & 0.69 & 0.7 & 0.77 & 0.76 & 0.76 & 0.76 & 0.76  \\ 
	Gauss\_5\_5\_5 & 0.74 & 0.71 & 0.69 & 0.66 & 0.74 & 0.75 & 0.74 & 0.74 & 0.74  \\ 
	Gauss\_5\_10\_5 & 0.78 & 0.74 & 0.68 & 0.58 & 0.73 & 0.72 & 0.74 & 0.74 & 0.73  \\ 
	Gauss\_5\_50\_5 & 0.65 & 0.6 & 0.62 & 0.57 & 0.62 & 0.62 & 0.63 & 0.63 & 0.62  \\ \hline
\end{tabular}

\caption{AMI For Algorithms}
\end{figure*}

\begin{figure*}[h]
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap   \\ \hline
	Soybean & 1.17 & 1.2 & 1.03 & 1.01 & 1.02 & 1.01 & 1.02  \\ 
	Iris & 1.08 & 1.37 & 1.02 & 1 & 1.01 & 1 & 1.01  \\ 
	Gauss\_5\_5\_5 & 1.07 & 1.44 & 1.03 & 1 & 1.03 & 1.01 & 1.03  \\ 
	Gauss\_5\_10\_5 & 1.08 & 1.39 & 1.02 & 1.01 & 1.02 & 1.01 & 1.02  \\ 
	Gauss\_5\_50\_5 & 1.09 & 1.27 & 1.03 & 1.02 & 1.03 & 1.01 & 1.03  \\ 
	pmed1 & 1.11 & 1.41 & 1.03 & 1.01 & 1.03 & 1.01 & 1.02  \\ \hline
\end{tabular}

\caption{Approximation Coeffiecients For Algorithms}
\end{figure*}

In terms of computation time, our proposed algorithms were much slower than $k$-means, the spectral algorithm, or $k$-medoids. In general, the multiswaps algorithm was around two orders of magnitude slower with a small multiplicative increase as $p$ increased, and the supermodular algorithm was around four orders of magnitude slower. This was probably due to the extra sampling the algorithm had to do.

@@

\section{Hierarchical Clustering Approximation Algorithms}
Suppose we have a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$. In the hierarchical setting, we will represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$. Therefore, for any $i \in S$, $a^{-1}(i)$ will be all points in the cluster centered at $i$. We define
\[ \mathrm{cost}(S,a) := \sum_{i \in X} d(i, a(i)) . \]

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ if 
\begin{enumerate}
\item $S_1 \subset S_2$, 
\item  $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, and 
\item $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
\end{enumerate}
 A solution to the hierarchical clustering problem is therefore a chain of nested solutions. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the algorithm.  
Consider any $k$-median algorithm $\mathcal{A}$, such as the algorithms above. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce  augmentation step that will nest these clusterings so that they fit into the hierarchical framework and only increase the costs by a bounded factor.

@@

We test this hierarchical framework for both the multiswap and supermodular local search algorithms above.

@@

For each data set and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information.

@@

For $k$-medoids in the non-logarithmic version and PAM in the logarithmic version, the hierarchical framework increased the computation time by three orders of magnitude, whereas this was decreased to two orders difference for $k$-medoids logarithmic version. Interestingly, the approximation factors did not seem to change between the randomized and deterministic versions for $k$-medoids. The approximation ratios seemed best as $n$ increased or as the dimension increased. In particular, for the Gaussian data in 7 dimensions, the approximation ratio was often around 3. Recall that this approximation measures the constructed $k$-median solution on every level against an optimal one so we expect the ratio to not come close to 1 in many cases.

@@

In the future, we hope to explore how linear programming based methods perform in this context as well and to implement our own agglomerative clustering so we can test it on the OR library instances. Lastly, when nesting solutions in the logarithmic framework, we could use agglomerative clustering to make smarter local merges.

>>>>

\begin{figure}[h]
\begin{tabular}{ | l | l | l | l | l | }
\hline
 Dataset Name & Opt Obj & $k$ & $n$ & dim \\ \hline
Soybean & 114 & 4 & 47 & 35 \\ 
Iris & 98 & 3 & 150 & 4 \\ 
Gauss\_5\_5\_5 & 195.4 & 5 & 388 & 5 \\ 
Gauss\_5\_10\_5 & 270 & 10 & 544 & 5 \\ 
Gaussian avg & 241 & 22 & 507 & 5 \\ 
pmed5 & 1355 & 33 & 100 & \  \\ 
pmed20 & 1789 & 133 & 400 & \  \\ 
% pmed40 & 5128 & 90 & 900 & \  \\ 
pmedian avg & 5535 & 46 & 460 & \  \\ \hline
 \end{tabular}
\caption{Some specifics about our datasets}
\end{figure}

Tests were conducted using Python 2.7.9 on a 64-bit Windows platform, on a 4 core Intel processor. Results are given in Figure 2  below with some results averaged due to space constraints.

\begin{figure*}[h]
\label{fig:flat_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Time & Time & Time & Time & Time & Time & Time & Time & Time \\ \hline
	&K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap  \\ \hline
	Soybean & 0.015 & 0.061 & 0.002 & 1.179 & 0.026 & 0.016 & 0.048 & 0.049 & 0.081 \\ 
	Iris & 0.013 & 0.023 & 0.003 & 9.788 & 0.125 & 0.102 & 0.212 & 0.229 & 0.414 \\ 
	Gauss\_5\_5\_5 & 0.025 & 0.056 & 0.032 & 75.262 & 1.159 & 1.506 & 2.188 & 2.382 & 3.71 \\ 
	Gauss\_5\_10\_5 & 0.054 & 0.095 & 0.109 & 134.137 & 6.834 & 5.407 & 9.243 & 9.428 & 15.96 \\ 
	%Gauss\_5\_50\_5 & 0.137 & 0.196 & 0.302 & 97.336 & 27.856 & 21.446 & 32.407 & 39.614 & 44.69 \\ 
	Gaussian avg & 0.073 & 0.128 & 0.182 & 144.455 & 13.289 & 13.021 & 18.749 & 18.696 & 24.592 \\ 
	pmed5 & 0.038 & 8.298 & 0.481 & 0.622 & 0.653 & 0.748 & 0.939 & \  & \  \\ 
	pmed20 & 1.795 & 174.951 & 52.208 & 44.979 & 68.554 & 68.231 & 79.288 & \  & \  \\ 
	% pmed40 & 3.527 & 849.852 & 251.259 & 264.867 & 250.412 & 391.738 & 350.235 & \  & \  \\ 
	pmedian avg & 0.807 & 203.308 & 40.386 & 40.347 & 48.264 & 56.76 & 56.19 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI & AMI  \\ \hline
	& K-means & Spectral & k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps + & 3-Swap   \\ \hline
	Soybean & 0.686 & 0.178 & 0.55 & 0.612 & 0.708 & 0.732 & 0.717 & 0.708 & 0.712 \\ 
	Iris & 0.748 & 0.757 & 0.69 & 0.7 & 0.772 & 0.760 & 0.763 & 0.757 & 0.761 \\ 
	Gauss\_5\_5\_5 & 0.741 & 0.713 & 0.687 & 0.657 & 0.742 & 0.749 & 0.741 & 0.744 & 0.737  \\ 
	Gauss\_5\_10\_5 & 0.783 & 0.736 & 0.677 & 0.575 & 0.729 & 0.725 & 0.737 & 0.737& 0.729  \\ 
	%Gauss\_5\_50\_5 & 0.646 & 0.596 & 0.621 & 0.570 & 0.622 & 0.619 & 0.627 & 0.631 & 0.625  \\ 
	Gaussian avg & 0.777 & 0.744 & 9.112 & 0.623 & 0.76 & 0.761 & 0.762 & 0.761 & 0.76 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | l | l | }
\hline
	Dataset Name & Approx & Approx & Approx & Approx & Approx & Approx & Approx  \\ \hline
	& k-medoids & Supermodular & PAM & 1-Swap + & 2-Swap & 2-swaps+ & 3-Swap   \\ \hline
	Soybean & 1.167 & 1.199 & 1.027 & 1.012 & 1.022 & 1.012 & 1.024 \\ 
	Iris & 1.081 & 1.375 & 1.016 & 1.004 & 1.013 & 1.004 & 1.009 \\ 
	Gauss\_5\_5\_5 & 1.070 & 1.436 & 1.030 & 1.002 & 1.027 & 1.005 & 1.026  \\ 
	Gauss\_5\_10\_5 & 1.080 & 1.389 & 1.017 & 1.011 & 1.018 & 1.009 & 1.017  \\ 
	%Gauss\_5\_50\_5 & 1.0887 & 1.272 & 1.030 & 1.020 & 1.027 & 1.014 & 1.033  \\ 
	Gaussian avg & 1.064 & 1.416 & 1.025 & 1.009 & 1.023 & 1.008 & 1.024 \\ 
	pmed5 & 1.373 & 1.681 & 1.06 & 1.026 & 1.071 & 1.033 & 1.051 \\ 
	pmed20 & 1.501 & 1.73 & 1.079 & 1.032 & 1.076 & 1.027 & 1.073 \\ 
	%pmed40 & 1.187 & 1.423 & 1.046 & 1.017 & 1.043 & 1.019 & 1.039 \\ 
	pmedian avg & 1.203 & 1.456 & 1.041 & 1.017 & 1.039 & 1.016 & 1.037 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), AMI measure, and Approximation Coeffiecients for Partitional Clustering Algorithms}
\end{figure*}

In terms of computation time, our proposed algorithms were much slower than $k$-means, the spectral algorithm, or $k$-medoids. In general, the multiswaps algorithm was around two orders of magnitude slower with a small multiplicative increase as $p$ increased, and the supermodular algorithm was around four orders of magnitude slower. This was probably due to the extra potential function sampling.

@@

\section{Hierarchical Clustering Approximation Algorithms}
Suppose we have a set $X$ and a distance metric $d(i,j)$ for all $i,j \in X$. In the hierarchical setting, we will represent a solution to the $k$-median problem as a pair $(S, a)$ where $S \subseteq X$ such that $|S| \leq k$ and $a$ is an assignment function that maps each $i \in X$ to a center in $S$. Therefore, for any $i \in S$, $a^{-1}(i)$ will be all points in the cluster centered at $i$. We define $\mathrm{cost}(S,a) := \sum_{i \in X} d(i, a(i))$.

Given solution pairs $(S_1, a_1)$ and $(S_2, a_2)$, we say that $(S_1, a_1)$ is \emph{nested} in $(S_2, a_2)$ if  (1) $S_1 \subset S_2$, (2)  $\forall  j \in X$, if $a_2(j) \in S_1$, then $a_1(j) = a_2(j)$, and (3) $\forall j, k \in X$, if $a_2(j) = a_2(k)$, then $a_1(j) = a_1(k)$.
 A solution to the hierarchical clustering problem is therefore a chain of nested solutions. However, we want this solution to stay close to the optimal $k$-median solutions when looking at any level of the algorithm.  
Consider any $k$-median algorithm $\mathcal{A}$, such as the algorithms above. After running $\mathcal{A}$ for all potential values of $k$, Lin et al.~\cite{Lin} introduce  augmentation step that will nest these clusterings into the hierarchical framework and only increase the costs by a bounded factor.

@@

@@

For each data set and each clustering method, we report the approximation ratio, computation time, and FScore measure.  FScore is a measure that looks at the clusters represented on each level and how they represent the true clustering in terms of precision and recall. See \cite{Larsen} for more information.  Results are in Figure 3.

\begin{figure*}[h]
\label{fig:hier_results}
\small
\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & Time & Time & Time & Time & Time & Time \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.0002 & 0.0002 & 0.2 & 0 & 0.2 & 0.2 \\ 
	Iris & 0.0023 & 0.003 & 3.2 & 0.2 & 3.8 & 5.1 \\ 
	Gauss\_5\_5\_5 &  & & & & & \\
	Gauss\_5\_10\_5 & & & & & & \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.0858 & 0.0853 & 354.5 & 7.5 & 369.9 & 423.3 \\ 
	pmed5 & 1.2 & 0.1 & 1.3 & 1.7 & \  & \  \\ 
	pmed25 & 279.3 & 5 & 306.7 & 312.5 & \  & \  \\ 
	pmedian avg & 87.4 & 2 & 94.5 & 101.6 & \  & \  \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | l | l | }
\hline
	   & F-Score & F-Score & F-Score & F-Score & F-Score & F-Score \\ \hline
	Dataset & Ave Linking & Ward Linking & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 0.6 & 0.5 & 0.8 & 0.6 & 0.7 & 0.6 \\ 
	Iris & 0.5 & 0.5 & 0.9 & 0.8 & 0.8 & 0.7 \\ 
	Gauss\_5\_5\_5 &  & & & & & \\
	Gauss\_5\_10\_5 &  & & & & & \\
	%Gauss\_5\_50\_5 &  & & & & & \\
	Gaussian Average & 0.4 & 0.4 & 0.6 & 0.6 & 0.6 & 0.6 \\ \hline
\end{tabular}
\vspace{0.2 cm}

\begin{tabular}{ | l | l | l | l | l | }
\hline
	   & Approx & Approx & Approx & Approx \\ \hline
	Dataset & K-med & K-med logn ran & k-med ran & PAM logn ran \\ \hline
	Soybean & 1.6 & 2.3 & 1.7 & 2 \\ 
	Iris & 2560631.5 & 2550279.6 & 2787312 & 1623132.1 \\ 
	Gauss\_5\_5\_5 &  & & &  \\
	Gauss\_5\_10\_5 &  & & &  \\
	%Gauss\_5\_50\_5 &  & & &  \\
	Gaussian Average & 5.2 & 7.3 & 5.5 & 4.7 \\ 
	pmed5 & 11.4 & 51.1 & 16.3 & 12 \\ 
	pmed25 & 6.6 & 13.4 & 5.4 & 13.2 \\ 
	pmedian avg & 10.3 & 19.8 & 10.5 & 8.6 \\ \hline
\end{tabular}

\caption{Average Computation Time (sec), FScore measure, and Approximation Coeffiecients for Hierarchical Clustering Algorithms}
\end{figure*}

@@

For $k$-medoids in the non-logarithmic version and PAM in the logarithmic version, the hierarchical framework increased the computation time by three orders of magnitude, whereas this was decreased to two orders difference for $k$-medoids logarithmic version. Interestingly, the approximation factors did not seem to change between the randomized and deterministic versions for $k$-medoids. The approximation ratios seemed best as $n$ increased or as the dimension increased. In particular, for the Gaussian data in 7 dimensions, the approximation ratio was often around 3. Recall that this approximation measures the constructed $k$-median solution on every level against an optimal one so we expect the ratio be higher than 1.  The high approximation ratio for the iris data set was due to the fact that the optimal solution for $k = n-1, n-2$ was almost equal to zero and our heuristic $k$-median algorithms only sometimes picked up on these solutions. The approximation factor on the lower levels was more consistent with the other results.

@@

In the future, we hope to explore how linear programming based methods perform in both the flat and hierarchical context as well and to implement our own agglomerative clustering to test on the OR library instances and help make smarter local merges in the nesting framework.
